$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in the Presence of Non-Functional Properties

### 2014-04-02 JournalToSubmit2013/intro.tex
sections 4 and 5

Our work proposes  the $\pi$-\textit{Service Oriented Development Method} ($\pi$-SODM)
to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements from the early stages of software development.
\pisodm is aligned with the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.  The technique proposes  the use of \textit{models} to specify a software system at different levels of abstraction. 
Models are organized according to the guidelines for structuring the specifications of software systems of the Model Driven Architecture (MDA)~\cite{miller} approach. 
The goals of  \pisodm are to:
%SOD-M does not provide support for the specification of non-functional requirements, such as
%security, reliability, and efficiency.
\begin{trivlist}
\item \textit{(i)} Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification;
\item \textit{(ii)} Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
\end{trivlist}

>>>>

Our work proposes  the $\pi$-\textit{Service Oriented Development Method} (\pisodm)
to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements from the early stages of software development.
\pisodm is aligned with the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.  The technique proposes  the use of \textit{models} to specify a software system at different levels of abstraction. 
Models are organized according to the guidelines for structuring the specifications of software systems of the Model Driven Architecture (MDA)~\cite{miller} approach. 
The goals of  \pisodm are to:
%SOD-M does not provide support for the specification of non-functional requirements, such as
%security, reliability, and efficiency.
\begin{trivlist}
\item \textit{(i)} Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification;
\item \textit{(ii)} Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
\end{trivlist}

### 2014-04-02 JournalToSubmit2013/modelingWithPISODM.tex
sections 4 and 5

This section presents \pisodm, an MDD-based methodology for building service compositions with non-functional requirements. 
\pisodm provides meta-models for modelling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}). 
Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation  of a set of models at the other levels of abstraction.
The refinement process is driven by transformation rules specified between the meta-models.
\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{figs/piSOD-M_process.pdf}
\caption{$\pi$SOD-M Overview.}
\label{fig:piSOD-M}
\end{figure}

@@

PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}. 
Figure~\ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
\begin{itemizedTrivlist}
\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}. 
A {\sc Compound Operation} is defined using an {\sc Operator}.
The language defines operators to denote guarded operations ($[C]S$); sequential ($\ . \ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetition.

\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
A brief description of these classes is given next.
\end{itemizedTrivlist}
%
\begin{figure}[t]
\centering
\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
\caption{$\pi$-{\sc Pews} Metamodel}
\label{fig:metamodel}
\end{figure}

Figure~\ref{fig:metamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.  
Each {\sc A-Policy} groups a set of ECA rules with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}.  
In this way, an {\em A-policy} represents a set of reactions to be possibly executed when one or several events are notified.

Given a $\pi$-SCM model of a specific service based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model. 
The following section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
%\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\section{Transformation rules}\label{sec:mmrules}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--

 $\pi$-SOD-M  proposes  horizontal transformation rules between $\pi$-use case to $\pi$-service process models,  and vertical transformation rules between $\pi$-service process to $\pi$-service composition  models; and $\pi$-service composition model to $\pi$-PEWS code. The following lines describe the particular purpose of these transformation rules that are intended to (semi)-automatize the construction of a service oriented application (see the $\pi$-SODM environment description in the next section).

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
\subsection{From $\pi$-UseCase to the $\pi$-ServiceProcess}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .

Given  (composite) $\pi$-use case model defining the functions of  the developed application and  associated constraints (representing business rules), the objective is to express them as a $\pi$-service process model describing the application logic and its associated non-functional properties.  The principle is to express a set of $\pi$-use cases (i.e., use cases with constraints)   in terms of  a business process consisting of actions related by a control flow and associated contracts specifying NFP. The PIM to PIM transformation rules specify how  to transform a $\pi$-use case model into  a $\pi$-service process model.
%(Table \ref{fig:transformationsUseCase-ServiceProcess}).

%\begin{figure}
%\centering{
%\includegraphics[width=0.96\textwidth]{figs/Table5}}
%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess transformation}
%\label{fig:transformationsUseCase-ServiceProcess}
%\end{figure}

As defined in the $\pi$-ServiceProcess  meta-model an {\sc Activity service}  consists of a composition of  entities of type {\sc Action}. 
Thus, every {\sf $\pi$-use case} is transformed into an {\sf Action} of the target $\pi$-service process model.  Every {\sf Extend} relationship identified in a $\pi$-use case model is transformed into a  {\sf Fork node}.
If the {\sf Extend} relationship concerns: 
\begin{itemize}
\item one {\sf use case} it is transformed into an {\sf Action} participating in the fork node in an alternative flow. 
\item several  {\sf use cases} are transformed as  different {\sf Actions} that belong to mutual alternative flows in the fork node.   
%{\sc Contracts}  are associated to  {\sc Actions}.
\end{itemize}

\begin{figure}
\centering{
\includegraphics[width=0.96\textwidth]{figs/35}}
\caption{ $\pi$-UseCase to $\pi$-ServiceProcess transformation rules}
\label{fig:transformationsUseCase-ServiceProcessRules}
\end{figure}

   A {\sf Constraint} of a (composite) {\sf Use Case}  is transformed into an  {\sf Assertion}.  The set of resulting assertions  are grouped in a {\sf Contract} in a $\pi$-service process model (figure  \ref{fig:transformationsUseCase-ServiceProcessRules}-a).
A constraint of a given type ({\sf Constraint type}) can have different transformations as stated in the following rules (figure   \ref{fig:transformationsUseCase-ServiceProcessRules}-b). Given a  a {\sf Constraint} of type:

\begin{itemize}
\item   {\sc Business} it is transformed into an {\sf Assertion} with no value attributes. %information like maxValue and minValue;
\item     {\sc Value}  with the  attribute {\sf  isExceptionalBehaviour}  set to false, is transformed into an {\sf Assertion}.
\item  {\sc Value }  with the attribute {\sf  isExceptionalBehaviour }  set to true, is transformed into  {\sf Exceptional behaviour}.
\end{itemize}
%Given the $\pi$-UseCase meta-model, to all {\sc Constraint} entity related with a {\sc Use Case}, there is a {\sc Contract} that compounds a set of {\sc Assertions} entity  and the {\sc Use Case} is refined in a service {\sc Action}.

 For transforming constraints of type {\sf Value Constraint }
  the designer must specify value thresholds associated to attributes as specified by  { constraint}s  in $\pi$-{use case}s. By default, value constraints are transformed into pre-conditions and business constraints are transformed into post-conditions.

% From {\sc Constraint} and {\sc Constraint Type} entities are generated detailed {\sc Contract} information, that are refined into {\sc Exceptional behaviours} and {\sc Assertions} entities . {\sc Constraint} is related with a {\sc Constraint type}, and there are different cases for the transformation of this concept. A Business type is transformed into an {\sc Assertion} . These {\sc  Assertion} information attributes are considered if the {\sc Constraint Type} is a Value type. This is semi-automatic, because there is not
%enough information in a $\pi$-UseCase model to run a complete automatic transformation.

A {\sf Requirement} and a {\sf Composite use case} in a  $\pi$-UseCase model are transformed respectively into a {\sf Service activity} and an {\sf Action} (see figures  \ref{fig:transformationsUseCase-ServiceProcessRules}-c  and d).

\begin{figure}
\centering{
\includegraphics[width=0.80\textwidth]{figs/37}}
\caption{ Extended Transformation Examples}
\label{fig:transformaton-examples}
\end{figure}

%The transformations for {\sc Extend} and {\sc Include} dependencies  are not as simple as the previous transformations (figures \ref{fig:transformaton-examples}-a).

The relationships of type  {\sc Extend} and {\sc Include}  determine the way the business process is expressed as a workflow.  The workflow generated for a relationship  «extends»  between  two use cases (figure  \ref{fig:transformationsUseCase-ServiceProcessRules}-a) is described in figure \ref{fig:transformaton-examples}-c).  The generated workflow consists of  a {\sf Fork node}, a {\sf Join node},  four entities  {\sc Control flow}, and also two entities of type {\sc Action}  (one for each {\sf Use case}).

%When there is more than one «extends» relationship among  use cases (figures \label{fig:transformationsUseCase-ServiceProcessRuless}-b), the transformation leads to  two more {\sc Control flow} entities for each  {\sc use case}, and an {\sf Action} for each  extended {\sf use case}.

\begin{figure}
\centering{
\includegraphics[width=0.96\textwidth]{figs/36}}
\caption{ $\pi$-UseCase to $\pi$-ServiceProcess Model Transformation Rules (2)}
\label{fig:transformationsUseCase-ServiceProcessRules}
\end{figure}

 {\sf Include} use case entities, are transformed into an {\sf Action} sequence, as shown in figure \ref{fig:transformationsUseCase-ServiceProcessRules}-b. A {\sc Use case} element is transformed into an {\sf Action}. A set of n {\sc Use case} is transformed into an  n-1 {\sf Object flow} elements. 
 %Each {\sf Control flow} is transformed into two {\sf Actions} associated by a relationship.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT1}
Considering the scenario example  the transformation of the "listen music" Use Case is transformed into a Service Action. This Action  represents a Spotify service function that can be invoked to play the music. For the "publish music" use case,  constraints are transformed in a set of assertions that are grouped in a Contract ({\sf "publishMusicContract"}) associated to the Action {\sf "publishMusic"}. 
The "download music" use case  includes the payment process to buy the music. Thus, these use cases  are transformed into {\sf Actions}, and a {\sf Service activity} that aggregates these {\sf Actions}.   It is transformed in a sequence flow in the $\pi$-service process model.
%(as shown in figure \ref{fig:transformation-example-include}). 
The same rule is applied for the "publish music" use case, which has two extended use cases, "public twitter" and "public Facebook" (figures \ref{fig:transformaton-examples}-b and d).
 \end{example}

%\begin{figure}
%\centering{
%\includegraphics[width=0.76\textwidth]{figs/38}}
%\caption{ Include Transformation Example}
%\label{fig:transformation-example-include}
%\end{figure}

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
\subsection{From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
%The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model. The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.   The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.

%As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}. 
%The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.

%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.

%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}. The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.

%
%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
%In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario. Thus in both cases the events are of type {\sf ActivityPrepared}. These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}. In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
%\begin{figure}[htpb]
%\centering{
%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
%\label{fig:p-scim}
%\end{figure}

%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
%\begin{figure}
%\centering{
%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
%\caption{Pews program implementing the "To Publish Music" application}
%\label{fig:pewsexpression}
%\end{figure}
%Table \ref{fig:ServiceProcess-ServiceComposition}  shows the transformation rules for transforming a .
As illustrated in Figure \ref{fig:ServiceProcess-ServiceComposition-Rules} the  principle of the transformation  of a  $\pi$-Service process model into a   $\pi$-ServiceComposition model is to group respectively  {\sf Contracts}  and {\sf Actions} into {\sf Policies} and {\sf Service activities}.

Each {\sf Assertion} of a {\sf Contract} in a $\pi$-service composition model is transformed into a {\sf Rule} in a $\pi$-service composition model. The set of {\sf Rules}  concerning the same NFP (e.g., safety, performance, and reliability) is grouped into a {\sf Policy} (see Figure \ref{fig:ServiceProcess-ServiceComposition-Rules}-a). Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule:Condition} attribute. If the {\sf Assertion} has a value type, the name and the attributes are transformed into a {\sf Variable} in the target model.  The {\sf Assertion: Aproperty} attribute can have different transformations, according to  the following rules:
\begin{itemize}
\item a {\sf Post-Condition} is transformed into a {\sf  Post}. 
\item a {\sf Precondition} is transformed into {\sf Pre}.
\item a {\sf Timerestriction} is transformed into {\sf Time}.
\end{itemize}

\begin{figure}
\centering{
\includegraphics[width=0.96\textwidth]{figs/ExceptionalRules}}
\caption{$\pi$-ServiceProcess to $\pi$-ServiceComposition Model Transformation Rules}
\label{fig:ServiceProcess-ServiceComposition-Rules}
\end{figure}

A {\sf Package}  in the $\pi$-use case model is transformed   into a {\sf Business Collaborator}. 
{\sf Actions}  of a $\pi$-service process model are transformed into an {\sf Action} in a $\pi$-service composition model, and every {\sf Service activity} of a $\pi$-service process model is transformed into a {\sf Service activity} in a $\pi$-service composition model (see Figure \ref{fig:ServiceProcess-ServiceComposition-Rules}-b). An {\sf Exceptional behaviour} entity is transformed into an {\sf Action} in a $\pi$-service composition model (see Figure \ref{fig:ServiceProcess-ServiceComposition-Rules}-c ), and every {\sf Non-functional attribute} associated to an element ({\sf Contract} and {\sf Non-functional requirement}) in a $\pi$-service process model becomes a {\sf Non-functional attribute} associated to the corresponding element {\sf Policy} in a $\pi$-service composition model.
 Finally, {\sf Actions} are grouped by  a  {\sf Business collaborator}.

%\begin{figure}
%\centering{
%\includegraphics[width=1\textwidth]{figs/ServiceProcess-ServiceComposition}}
%\caption{ Transformation Rules: From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
%\label{fig:ServiceProcess-ServiceComposition}
%\end{figure}

%As the $\pi$-ServiceComposition model refines the $\pi$-ServiceProcess concepts at PIM level, a service previously defined as actions (source model) is refined as composition of those actions (target model) that are necessary to represent a business service, identifying who are the partners involved in the realization ({\sc Business collaborators}). 
%In addition, $\pi$-SOD-M defines a platform specific model based on web services composition. This model is explicitly indicates those actions which are (or will be, if not yet implemented) supported by web services.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}
Considering the scenario example, 
%the {\sf Action}s update music {\sc Contract} is transformed is a {\sc Policy} with its {\sc Rules}. All contract {\sc Assertions} are transformed in RULE and its attributes, e.g. the login and password verification. 
the "securityLoginPolicy" consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
%The {\sc Non- functional requirement} information will be used to the {\sc Policy} generation comes from the initial use case model. Also 
The {\sf Business Collaborator} Facebook and Spotify information come from entity of type {\sc Package}  in the $\pi$-use case model. 
%All {\sc Contracts} of the same {\sc Non- functional requirement} are composed in a {\sc Policy}.
\end{example}

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
\subsection{From $\pi$-ServiceComposition to $\pi$-PEWS}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .

%Table \ref{fig:ServiceComposition-Pews-Rules} 
This section describes the PIM to PSM transformations from a $\pi$-Service composition model to a $\pi$-PEWS program. We propose two groups of rules: those that transform services composition entities of a $\pi$-Service composition model into $\pi$-PEWS model; and those that transform rules grouped by policies into A-policy types.

A named action of the $\pi$-service composition model represented by  {\sf Action} and {\sf Action:name} is transformed to an  {\sf Operation} with a corresponding attribute name {\sf Operation:name}. A  named service activity represented by the elements {\sf ServiceActivity}  and  {\sf ServiceActivity:name} of the $\pi$-service composition model, are  transformed into a named operation of the $\pi$-{Pews} represented by the entities  {\sf CompositeOperation} and {\sf CompositeOperation:name}. When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-service composition model the corresponding transformations, according to the PEWS operators presented above:
\begin{itemize}
\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
 \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
\end{itemize}

%\begin{figure}
%\centering{
%\includegraphics[width=0.96\textwidth]{figs/Table7}}
%\caption{Transformation Rules: From $\pi$-ServiceComposition to $\pi$-PEWS}
%\label{fig:ServiceComposition-Pews-Rules}
%\end{figure}

The A-policies defined for the entities of a $\pi$-service composition model are transformed into {\sf A-policy} entities, named according to the names expressed in the source model. The transformation of the rules expressed in a $\pi$-service composition is guided by the event types associated to these rules. The variables associated to an A-policy expressed in a $\pi$-service composition model as {\sf $<$Variable:name, Variable:type$>$} are transformed into entities  {\sf Variable} with attributes {\sf Name} and {\sf Type} directly specified from the elements {\sf Variable:name} and {\sf Variable:type} of a $\pi$-service composition model.

%As shown in Table \ref{fig:ServiceComposition-Pews-Rules}, 
For an event of type Pre the corresponding transformed rule is a {\sf Precondition}; for an event of type Post the corresponding transformed rule is a {\sf Post- condition}; finally, for an event of type TimeRestriction the corresponding transformed rule is a {\sf Time}. The condition expression of a rule in a $\pi$-service composition ({\sf Rule:condition}) is transformed into a  {\sf Condition:expression} where the attributes of the expression are transformed into  entities {\sf Attribute}.

Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$- service composition model  of our scenario example.

\begin{figure}
\centering{
\includegraphics[width=0.96\textwidth]{figs/31-31}}
\caption{$\pi$-PEWS Specific and Contract Representation}
\label{fig:Specific-Contract-Representation}
\end{figure}

>>>>

This section presents \pisodm, an MDD-based methodology for building service compositions with non-functional requirements. 
\pisodm provides meta-models for modelling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}). 
Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
The refinement process is driven by transformation rules specified between the meta-models.

The next subsections present the meta-models from higher to lower levels of abstraction.
The presentation is exemplified by a running example.
The transformations between models are presented in section~\ref{sec:mmrules}.

\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{figs/piSOD-M_process.pdf}
\caption{\pisodm Overview.}
\label{fig:piSOD-M}
\end{figure}

@@

PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}. 
Figure~\ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
\begin{itemizedTrivlist}
\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}. 
A {\sc Compound Operation} is defined using an {\sc Operator}.
The language defines operators to denote guarded operations ($[C]S$); sequential ($\ . \ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) repetition.

\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
A brief description of these classes is given next.
\end{itemizedTrivlist}
%
\begin{figure}[t]
\centering
\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
\caption{$\pi$-{\sc Pews} Metamodel.}
\label{fig:PPEWSmetamodel}
\end{figure}

Figure~\ref{fig:PPEWSmetamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.  
Each {\sc A-Policy} groups a set of ECA rules with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}.  
In this way, an {\em A-policy} represents a set of reactions to be possibly executed when one or several events are notified.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic5}
Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$-service composition model of our scenario example.
Notice that the code contains namespaces and definitions (obtained by using information about the business collaborators of the $\pi$-SCM model), a workflow expression \textit{(Path)} containing operation calls and contracts (derived from the \textit{A-Policies}).
\end{example}

\begin{figure}
\centering{
\includegraphics[width=0.96\textwidth]{figs/31-31}}
\caption{$\pi$-PEWS Specific and Contract Representation. {\color{red}\sc Placido: Please change this figure\dots}}
\label{fig:Specific-Contract-Representation}
\end{figure}
